---
title: 'Project 2: Dropout Rates in Texas Public Schools from 1998 to 2019'
author: "Thomas Smith"
date: "4/21/2021"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Project Overview

For my second project, I wanted to explore dropout rates in Texas public schools. As a former high school math teacher, I worked with countless students, each with their own background, heritage and labels. A lot of those students struggled in high school, but luckily very few of them dropped out. 

According to Texas Education Agency (TEA), there are a multitude of reasons for dropping out of high school. Here are some common reasons:

- Poor attendance (attrition)
- Entered into alternative GED (non-diploma) program
- Not enough credits earned
- Didn't pass Standardized Exams (STAAR)
- Age being above the maximum limit
- Expulsion
- Incarceration
- Being homeless
- Deceased 

I decided to focus on three main factors: Grade level, Race/Ethnicity, and Student Characteristics.

Student characteristics include: Special Education, 504, Gifted and Talented (GT), At Risk, English Language Learner (ELL), dyslexic, ect.

I scraped data from [TEA's Website](https://tea.texas.gov/) from 1998-2019 and analyzed the data. Here are my results. 

***

```{r dropout, echo=FALSE}
library(shiny)
library(ggplot2)
#Load the Library
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total1.RData")
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total4.RData")
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total5.RData")
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total8.RData")
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total7.RData")
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/tot_Afr.RData")
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/tot_His.RData")
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/tot_Wh.RData")



# Define UI for application that draws a histogram
ui<-navbarPage("Exploring Dropout Rates By:",
           tabPanel("Grade Level",
                    
                    sidebarLayout(
                        #Inputs: Select which inputs from the data we want to display
                        sidebarPanel(
                          br(),
                          br(),
                          br(),
                            #Select variable for y-axis
                            selectInput(inputId = "a", 
                                        label = "Y-axis:",
                                        choices = c("Female","Female_Percentage","Male","Male_Percentage","State","State_Percentage"), 
                                        selected = "Female"),
                            #Select X-axis variables
                            selectInput(inputId = "b", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade"), 
                                        selected = "School_Year"),
                            #Select Grade Level
                            selectInput(inputId = "c", 
                                        label = "Color:",
                                        choices = c("Grade","School_Year"),
                                        selected = "Grade"),
                            hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            
                            selectInput(inputId = "d", 
                                        label = "Y-Axis:",
                                        choices = c("Fem_Grade_7","Fem_Grade_8","Fem_Grade_9","Fem_Grade_10","Fem_Grade_11","Fem_Grade_12"),
                                        selected = "Fem_Grade_7"),
                        
                            
                            selectInput(inputId = "e", 
                                        label = "X-axis:",
                                        choices = c("School_Year"), 
                                        selected = "School_Year"),
                            
                            # Select Colors
                            selectInput(inputId = "f", 
                                        label = "Color",
                                        choices =  c("red","green","blue","yellow","black"), 
                                        selected = "black"),
                            
                            hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            
                            selectInput(inputId = "g", 
                                        label = "Y-axis:",
                                        choices = c("Male_Grade_7","Male_Grade_8","Male_Grade_9","Male_Grade_10","Male_Grade_11","Male_Grade_12"),
                                        selected = "Male_Grade_7"),
                        
                            
                            selectInput(inputId = "h", 
                                        label = "X-axis:",
                                        choices = c("School_Year"), 
                                        selected = "School_Year"),
                            
                            # Select Colors
                            selectInput(inputId = "i", 
                                        label = "Color",
                                        choices =  c("red","green","blue","yellow","black"), 
                                        selected = "black")
                        ),
                        
                        
                        #Output: Type of plot
                        mainPanel(
                            titlePanel("Dropout Rates from 1998-2019"),
                            plotOutput(outputId = "Graphic"),
                            hr(),
                            titlePanel("Female Dropout Rate by Grade Level"),
                            plotOutput(outputId="Barplot"),
                            hr(),
                            titlePanel("Male Dropout Rate by Grade Level"),
                            plotOutput(outputId="Barplot1"),
                             hr(),
                            br(),
                            br(),
                            br(),
                            
                            titlePanel("Key Takeaways"),
                            br(),
                            titlePanel("Female Vs Male Total Dropouts from 1998-2019 (Excluding 2002-03)"),
                            plotOutput(outputId="Boxplot"),
                            p("Comparing the total number of dropouts each year from 1998-2003, overall, males in grades 7-12 had higher dropout totals than females. "),
                              hr(),
                            titlePanel("Female Vs Male Dropout Rate from 1998-2019 by Grade Level"),
                            plotOutput(outputId="Boxplot1"),
                          
                            titlePanel("Female Vs Male Dropout % by Grade Level"),
                            plotOutput(outputId="Boxplot2"),
                            p("Comparing male vs female dropout rates by grade level, as a whole, we can see that males had higher dropout rates in grades 7-12. Males had the highest number of dropouts in grade 9, but the highest percentage of dropouts occured in grade 12. Females had the highest number and percentage of dropouts in grade 12."),
                                hr(),
                            titlePanel("Female Vs. Male Dropouts Grades 7-12 by Year"),
                            plotOutput(outputId="Boxplot3"),
                            
                            titlePanel("Female Vs. Male Dropout % Grades 7-12 by Year"),
                            plotOutput(outputId="Boxplot4"),
                            p("Comparing male vs female dropout rates by school year, we can see that,as a whole, males had a higher rate of dropouts each year. The school year 2006-07 had the highest rate and percentage of female and male dropouts, while 2003-04 and 2004-05 had the lowest rate and percentage."),
                            

                        )

                    )
                ),

           tabPanel("Ethnicity/Race",
                    sidebarLayout(
                        #Inputs: Select which inputs from the data we want to display
                        sidebarPanel(
                          br(),
                          
                          br(),
                          
                          br(),
                          
                          br(),
                            #Select variable for y-axis
                            selectInput(inputId = "j", 
                                        label = "Y-axis:",
                                        choices = c("Female_Dropouts","Female_Dropouts_Percentage","Male_Dropouts","Male_Dropouts_Percentage","Annual_Female_Dropout_Rate","Annual_Male_Dropout_Rate"), 
                                        selected = "Female_Dropouts"),
                            #Select X-axis variables
                            selectInput(inputId = "k", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "School_Year"),
                            #Select Grade Level
                            selectInput(inputId = "l", 
                                        label = "Color:",
                                        choices = c("Race","Grade_Level"),
                                        selected = "Race"),
                            hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                          
                            br(),
                          
                          selectInput(inputId = "m", 
                                        label = "Y-axis:",
                                        choices = c("Female_Dropouts","Female_Dropouts_Percentage","Male_Dropouts","Male_Dropouts_Percentage","Annual_Female_Dropout_Rate","Annual_Male_Dropout_Rate"),
                                        selected = "Female_Dropouts"),
                        
                            
                            selectInput(inputId = "n", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "School_Year"),
                            
                           hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                          
                          selectInput(inputId = "o", 
                                        label = "Y-axis:",
                                        choices = c("Female_Dropouts","Female_Dropouts_Percentage","Male_Dropouts","Male_Dropouts_Percentage","Annual_Female_Dropout_Rate","Annual_Male_Dropout_Rate"),
                                        selected = "Female_Dropouts"),
                        
                            
                            selectInput(inputId = "p", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "School_Year"),
                          
                          selectInput(inputId = "q", 
                                        label = "Label:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "Grade_Level"),
                           hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                          
                            br(),
                          
                          
                          selectInput(inputId = "r", 
                                        label = "Y-axis:",
                                        choices = c("Female_Dropouts","Female_Dropouts_Percentage","Male_Dropouts","Male_Dropouts_Percentage","Annual_Female_Dropout_Rate","Annual_Male_Dropout_Rate"),
                                        selected = "Female_Dropouts"),
                        
                            
                            selectInput(inputId = "s", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "School_Year"),
                           hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                        
                          
                          selectInput(inputId = "t", 
                                        label = "Y-axis:",
                                        choices = c("Female_Dropouts","Female_Dropouts_Percentage","Male_Dropouts","Male_Dropouts_Percentage","Annual_Female_Dropout_Rate","Annual_Male_Dropout_Rate"),
                                        selected = "Female_Dropouts"),
                        
                            
                            selectInput(inputId = "u", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "School_Year"),
                          
                          selectInput(inputId = "v", 
                                        label = "Label:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "Grade_Level"),
      
                           hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                          
                          selectInput(inputId = "w", 
                                        label = "Y-axis:",
                                        choices = c("Female_Dropouts","Female_Dropouts_Percentage","Male_Dropouts","Male_Dropouts_Percentage","Annual_Female_Dropout_Rate","Annual_Male_Dropout_Rate"),
                                        selected = "Female_Dropouts"),
                        
                            
                            selectInput(inputId = "x", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "School_Year"),
                           hr(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                            
                            br(),
                          
                          
                          selectInput(inputId = "y", 
                                        label = "Y-axis:",
                                        choices = c("Female_Dropouts","Female_Dropouts_Percentage","Male_Dropouts","Male_Dropouts_Percentage","Annual_Female_Dropout_Rate","Annual_Male_Dropout_Rate"),
                                        selected = "Female_Dropouts"),
                        
                            
                            selectInput(inputId = "z", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "School_Year"),
                          
                          selectInput(inputId = "a1", 
                                        label = "Label:",
                                        choices = c("School_Year","Grade_Level"), 
                                        selected = "Grade_Level")
                            
                        ),
                        
                        #Output: Type of plot
                        mainPanel(
                            titlePanel("Dropout Rates from 2002-2019 By Race/Ethnicity"),
                            plotOutput(outputId = "Graphic2"),
                            hr(),
                            titlePanel("African American Dropout Rates from 2002-2019"),
                            plotOutput(outputId = "Ethnicity_Barplot"),
                            plotOutput(outputId = "Ethnicity_Labelplot"),
                            hr(),
                            titlePanel("Hispanic Dropout Rates from 2002-2019"),
                            plotOutput(outputId = "Ethnicity_Barplot1"),
                            plotOutput(outputId = "Ethnicity_Labelplot1"),
                            hr(),
                            titlePanel("White Dropout Rates from 2002-2019"),
                            plotOutput(outputId = "Ethnicity_Barplot2"),
                            plotOutput(outputId = "Ethnicity_Labelplot2"),
                            
                        )
                    )
           ),
           
           tabPanel("Student Characteristics",
                    sidebarLayout(
                        #Inputs: Select which inputs from the data we want to display
                        sidebarPanel(
                            #Select variable for y-axis
                            selectInput(inputId = "b1", 
                                        label = "Y-axis:",
                                        choices = c("Students","Students_Percentage","Dropouts","Dropouts_Percentage","Annual_Dropout_Rate","Male_Dropouts_Percentage"), 
                                        selected = "Dropouts"),
                            #Select X-axis variables
                            selectInput(inputId = "c1", 
                                        label = "X-axis:",
                                        choices = c("School_Year","Groups"), 
                                        selected = "School_Year"),
                            #Select Grade Level
                            selectInput(inputId = "d1", 
                                        label = "Color:",
                                        choices = c("Groups","School_Year"),
                                        selected = "Groups")
                        ),
                        
                        #Output: Type of plot
                        mainPanel(
                            plotOutput(outputId = "Graphic3"),
                        )
                    )
           )
)

server<-function(input, output) {
    output$Graphic <- renderPlot({
        # draw the histogram with the specified number of bins
        ggplot(total1, aes_string(x=input$b, y=input$a,col=input$c)) + geom_jitter()
      #Notice the difference between the ggplots
    })
    
    output$Barplot <- renderPlot({
        ggplot(total4, aes_string(x=input$e, y=input$d)) + geom_col(color=input$f)
    })
    
    output$Barplot1 <- renderPlot({
        ggplot(total4, aes_string(x=input$h, y=input$g)) + geom_col(color=input$i)
    })
    
    output$Boxplot <- renderPlot({
      mycolor<-c("red","blue")

#Boxplot Comparing Male/Female Dropout Rate Totals from 1998-2019 (Excluding 2002-03)
boxplot(total3[,c(3,5)],col=mycolor)
    })
    
     output$Boxplot1 <- renderPlot({
       boxplot(total4[,c(3,5,12,14,20,22,28,30,36,38,44,46)],xlab="Grade 7-12",ylab="Dropouts",col=mycolor)
legend("topleft",legend = c("Female", "Male"),col = mycolor,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))
    })
     
     output$Boxplot2 <- renderPlot({
       boxplot(total4[,c(4,6,13,15,21,23,29,31,37,39,45,47)],xlab="Grade 7-12",ylab="Dropouts",col=mycolor)
legend("topleft",legend = c("Female", "Male"),col = mycolor,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))
    })
     
     output$Boxplot3 <- renderPlot({
       mycolor1<-rep(c("Orange","Green"),21)

#Boxplot of Female Vs Male Dropouts from grade 7-12 by Year: 1998-2019
boxplot(total2[,c(3,5,11,13,19,21,27,29,35,37,43,45,51,53,59,61,67,69,75,77,83,85,91,93,99,101,107,109,115,117,123,125,131,133,139,141,147,149,155,157,163,165)],xlab="School Year: 1998-2019",ylab="Dropout Rate",col=mycolor1)
legend("topleft",legend = c("Female", "Male"),col = mycolor1,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))
    })
     
     output$Boxplot4 <- renderPlot({
       #Boxplot of Female Vs Male Dropout % from grade 7-12 by Year: 1998-2019
boxplot(total2[,c(4,6,12,14,20,22,28,30,36,38,44,46,52,54,60,62,68,70,76,78,84,86,92,94,100,102,108,110,116,118,124,126,132,134,140,142,148,150,156,158,164,166)],xlab="School Year: 1998-2019",ylab="Dropout Rate",col=mycolor1)

legend("topleft",legend = c("Female", "Male"),col = mycolor1,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))
       
     })
     
    output$Graphic2 <- renderPlot({
        # draw the histogram with the specified number of bins
        ggplot(total5, aes_string(x=input$k, y=input$j,col=input$l)) + geom_jitter()
    })
    
     output$Ethnicity_Barplot <- renderPlot({
        ggplot(tot_Afr, aes_string(x=input$n, y=input$m)) + geom_col()
    })
     
     output$Ethnicity_Labelplot <- renderPlot({
        ggplot(tot_Afr, aes_string(x=input$p, y=input$o,label=input$q)) + geom_label()
    })
    
    output$Ethnicity_Barplot1 <- renderPlot({
        ggplot(tot_His, aes_string(x=input$s, y=input$r)) + geom_col()
    })
    
    output$Ethnicity_Labelplot1 <- renderPlot({
        ggplot(tot_Afr, aes_string(x=input$u, y=input$t,label=input$v)) + geom_label()
    })
    
    output$Ethnicity_Barplot2 <- renderPlot({
        ggplot(tot_Wh, aes_string(x=input$x, y=input$w)) + geom_col()
    })
    
    output$Ethnicity_Labelplot2 <- renderPlot({
        ggplot(tot_Afr, aes_string(x=input$z, y=input$y,label=input$a1)) + geom_label()
    })
    output$Graphic3 <- renderPlot({
        # draw the histogram with the specified number of bins
        ggplot(total8, aes_string(x=input$c1, y=input$b1,col=input$d1)) + geom_jitter() 
    })
}



# Run the application 
shinyApp(ui = ui, server = server)

```

# Key Takeaways from Results

### Dropouts by Grade Level

```{r dropout1, echo=FALSE}
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total3.RData")

mycolor<-c("red","blue")

#Boxplot Comparing Male/Female Dropout Rate Totals from 1998-2019 (Excluding 2002-03)
boxplot(total3[,c(3,5)],col=mycolor,main="Female Vs Male Annual Dropout Totals from 1998-2019(Excludes 02-03)")
```

#### Comparing the total number of dropouts each year from 1998-2003, overall, males in grades 7-12 had higher dropout totals than females. 

```{r dropout2, echo=FALSE}
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total4.RData")

mycolor<-c("red","blue")

#Boxplot Comparing Male/Female Dropout Rate by Grade-Level
boxplot(total4[,c(3,5,12,14,20,22,28,30,36,38,44,46)],main="Female Vs Male Dropout Rate from 1998-2019 by Grade Level",xlab="Grade 7-12",ylab="Dropouts",col=mycolor)
legend("topleft",legend = c("Female", "Male"),col = mycolor,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))

boxplot(total4[,c(4,6,13,15,21,23,29,31,37,39,45,47)],main="Female Vs Male Dropout % by Grade Level",xlab="Grade 7-12",ylab="Dropouts",col=mycolor)
legend("topleft",legend = c("Female", "Male"),col = mycolor,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))

```

#### Comparing male vs female dropout rates by grade level, as a whole, we can see that males had higher dropout rates in grades 7-12. Males had the highest number of dropouts in grade 9, but the highest percentage of dropouts occured in grade 12. Females had the highest number and percentage of dropouts in grade 12. 

```{r dropout3, echo=FALSE}
load("C:/Users/12108/OneDrive/Desktop/UTSA/Spring 2021/R Project/R_Projects/Project 2/total2.RData")

mycolor1<-rep(c("Orange","Green"),21)

#Boxplot of Female Vs Male Dropouts from grade 7-12 by Year: 1998-2019
boxplot(total2[,c(3,5,11,13,19,21,27,29,35,37,43,45,51,53,59,61,67,69,75,77,83,85,91,93,99,101,107,109,115,117,123,125,131,133,139,141,147,149,155,157,163,165)],main="Female Vs. Male Dropouts Grades 7-12 by Year",xlab="School Year: 1998-2019",ylab="Dropout Rate",col=mycolor1)

legend("topleft",legend = c("Female", "Male"),col = mycolor1,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))


#Boxplot of Female Vs Male Dropout % from grade 7-12 by Year: 1998-2019
boxplot(total2[,c(4,6,12,14,20,22,28,30,36,38,44,46,52,54,60,62,68,70,76,78,84,86,92,94,100,102,108,110,116,118,124,126,132,134,140,142,148,150,156,158,164,166)],main="Female Vs. Male Dropout % Grades 7-12 by Year",xlab="School Year: 1998-2019",ylab="Dropout Rate",col=mycolor1)

legend("topleft",legend = c("Female", "Male"),col = mycolor1,pch = 19,bty = "n",pt.cex = 1,cex = 0.7,text.col = "black",horiz = F ,inset = c(0.05, 0.05))

```

#### Comparing male vs female dropout rates by school year, we can see that overall males had a higher rate of dropouts each year. The year 2006-07 had the highest rate and percentage of female and male dropouts, while the year 2003-04 and 2004-05 had the lowest rate and percentage. 